Financialisation of Commodity Markets

### 2025-07-04 docs/paper.qmd
Update paper online version

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: the Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear

  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

>>>>

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: the Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---
